---
title: 'Using R in Teaching from *Network Science*'
author: "Amir Barghi, Department of Mathematics and Statistics, Saint Michael's College"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

# Chap. 5: The Barabasi-Albert Model

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# configuring figure size
options(repr.plot.width = 6, repr.plot.height = 4)
```

# Loading Packages

```{r}
# uncomment the following lines if you have not installed qdread/forestscaling; for more information, see https://github.com/qdread/forestscaling/

#install.packages('remotes')
#remotes::install_github('qdread/forestscaling')
```

```{r, message = FALSE}
library(tidyverse)

library(igraph)

library(igraphdata)

library(ggraph)

library(e1071)

library(forestscaling)

library(glue)

library(latex2exp)
```

# 5.3 The Barabasi-Albert Model 

```{r}
set.seed(42)
# the order of the graph
n <- 10
# the number of edges to be added at each step
for (m in c(2, 4, 8, 16, 32)) {
  
  g <- sample_pa(n, m = m, directed = FALSE, algorithm = 'psumtree')

  plot(g, 
       layout = layout_in_circle, 
       vertex.size = 2, 
       vertex.label = NA, 
       edge.lty = 3,
       main = 'An Example of Barabasi-Albert Model',
       sub = glue('with ', n, ' Nodes and ', m, ' Edges Added at Each Step'))
  
  writeLines('\n')

}
```

```{r}
set.seed(42)

n <- 100

for (m in c(2, 4, 8, 16, 32)) {
  
 g <- sample_pa(n, m = m, directed = FALSE, algorithm = 'psumtree')

plot(g, 
     layout = layout_in_circle, 
     vertex.size = 2, 
     vertex.label = NA, 
     edge.lty = 3,
     main = 'An Example of Barabasi-Albert Model',
     sub = glue('with ', n, ' Nodes and ', m, ' Edges Added at Each Step'))
 
 writeLines('\n')
 
}
```

```{r}
set.seed(42)

n <- 100

DF <- NULL

for (m in c(2, 4, 8, 16, 32)) {
  
  g <- sample_pa(n, m = m, directed = FALSE, algorithm = 'psumtree')
    
  print(ggraph(g, layout = 'circle') + 
          geom_edge_fan(edge_linetype = 3, color = 'dark blue', alpha = 0.25) +
          geom_node_point(color = 'dark red', size = 2, alpha = 0.75) + 
          theme_graph(base_family = 'Helvetica') +
          labs(title = 'An Example of Barabasi-Albert Model',
               subtitle = glue('with ', n, ' Nodes and ', m, ' Edges Added at Each Step')))
  
  writeLines('\n')

  suppressMessages(df <- bind_cols(enframe(eccentricity(g)), 
                                   enframe(betweenness(g)), 
                                   enframe(degree(g)), 
                                   enframe(transitivity(g, type = c('local')))))
  
  df <- df %>% select(name...1, value...2, value...4, value...6, value...8)

  names(df) <- c('name', 'eccentricity', 'betweenness', 'degree', 'clustering')
  
  DF2 <- df %>% 
    summarize(m = m,
              avg_deg = mean(degree), 
              delta = max(degree), 
              diam = max(eccentricity), 
              radius = min(eccentricity), 
              avg_cc = mean(clustering, na.rm = TRUE), 
              exponent = fit_power_law(degree, implementation = 'plfit')$alpha)
  
  DF <- rbind(DF, DF2)
}

DF
```

```{r}
set.seed(42)

n <- 1000

DF <- NULL

for (m in c(2, 4, 8, 16, 32)) {
  
  g <- sample_pa(n, m = m, directed = FALSE, algorithm = 'psumtree')
    
  print(ggraph(g, layout = 'circle') + 
        geom_edge_fan(edge_linetype = 3, color = 'dark blue', alpha = 0.25) + 
        geom_node_point(color = 'dark red', size = 2, alpha = 0.75) + 
        theme_graph(base_family = 'Helvetica') +
        labs(title = 'An Example of Barabasi-Albert Model',
             subtitle = glue('with ', n, ' Nodes and ', m, ' Edges Added at Each Step')))
  
  writeLines('\n')

  suppressMessages(df <- bind_cols(enframe(eccentricity(g)), 
                                   enframe(betweenness(g)), 
                                   enframe(degree(g)), 
                                   enframe(transitivity(g, type = c('local')))))
  
  df <- df %>% select(name...1, value...2, value...4, value...6, value...8)

  names(df) <- c('name', 'eccentricity', 'betweenness', 'degree', 'clustering')
  
  DF2 <- df %>% 
    summarize(m = m,
              avg_deg = mean(degree), 
              delta = max(degree), 
              diam = max(eccentricity),
              radius = min(eccentricity),
              avg_cc = mean(clustering, na.rm = TRUE), 
              exponent = fit_power_law(degree, implementation = 'plfit')$alpha)
  
  DF <- rbind(DF, DF2)
}

DF
```

```{r}
set.seed(42)

n <- 1000

for (m in c(2, 4, 8, 16, 32)) {
  
  g <- sample_pa(n, m = m, directed = FALSE, algorithm = 'psumtree')

  df <- enframe(degree(g))
  
  df <- df %>% select(name, value)

  names(df) <- c('name', 'degree')
  
  exponent <- fit_power_law(df$degree, implementation = 'plfit')$alpha
  
  df2 <- logbin(df$degree, n = 100)
  
  total2 <- sum(df2$bin_value)

  print(df2 %>% 
          filter(bin_count != 0) %>% 
          ggplot(aes(x = bin_midpoint, y = bin_value / total2)) + 
          geom_point(color = 'blue', size = .5) + 
          geom_line(data = df, 
                    aes(x = degree, y = 2 * m ** (exponent - 1) * (degree) ** (-exponent)), 
                    color = 'red') + 
          geom_line(data = df, 
                    aes(x = degree, y = 2 * m ** 2 * (degree) **(-3)), 
                    color = 'yellow') + 
          scale_x_log10() + 
          scale_y_log10() +
          labs(y = TeX('$p_k$')) +
          labs(title = 'Log-Log Scale, Linear Binning for Degree Distribution',
               subtitle = glue('with Fitted Power Law Curves (Eq. 5.9: Red; Eq. 5.15: Yellow) in a Barabasi-Albert Graph \nwith ', n, ' Nodes and ',  m, ' Edges Added at Each Step')))
  
  writeLines('\n')

}
```

# 5.5 Degree Distribution

```{r}
set.seed(42)

n <- 100

for (power in c(0, 0.5, 1, 1.5, 2, 2.5, 3)) {
  
  g <- sample_pa(n, power = power, directed = FALSE, algorithm = 'psumtree')

  plot(g, 
     layout = layout_in_circle, 
     vertex.size = 2, 
     vertex.label = NA, 
     edge.lty = 3,
     main = 'An Example of Barabasi-Albert Model',
     sub = glue('with ', n, ' Nodes and the Power of Preferential Attachment of ', power))
  
  writeLines('\n')

}
```

```{r}
set.seed(42)

n <- 100

DF <- NULL
# the power of the preferential attachment
for (power in c(0, 0.5, 1, 1.5, 2, 2.5, 3))  {
  
  m <- 8
  
  g <- sample_pa(n, m = m, power = power, directed = FALSE, algorithm = 'psumtree')
    
 print(ggraph(g, layout = 'circle') + 
          geom_edge_fan(edge_linetype = 3, color = 'dark blue', alpha = 0.25) + 
          geom_node_point(color = 'dark red', size = 2, alpha = 0.75) + 
          theme_graph(base_family = 'Helvetica') +
          labs(title = 'An Example of Barabasi-Albert Model',
               subtitle = glue('with ', n, ' Nodes, ', m, ' Edges Added at Each Step, \nand the Power of Preferential Attachment of ', power)))
  
  writeLines('\n')

  suppressMessages(df <- bind_cols(enframe(eccentricity(g)), 
                                   enframe(betweenness(g)), 
                                   enframe(degree(g)), 
                                   enframe(transitivity(g, type = c('local')))))
  
  df <- df %>% select(name...1, value...2, value...4, value...6, value...8)

  names(df) <- c('name', 'eccentricity', 'betweenness', 'degree', 'clustering')
  
  DF2 <- df %>% 
    summarize(power = power,
              avg_deg = mean(degree), 
              delta = max(degree), 
              diam = max(eccentricity),
              radius = min(eccentricity),
              avg_cc = mean(clustering, na.rm = TRUE), 
              exponent = fit_power_law(degree, implementation = 'plfit')$alpha)
  
  DF <- rbind(DF, DF2)
  
}

DF
```

```{r}
set.seed(42)

n <- 500

DF <- NULL

for (power in c(0, 0.5, 1, 1.5, 2, 2.5, 3)) {
  
  m <- 8
  
  g <- sample_pa(n, m = m , power = power, directed = FALSE, algorithm = 'psumtree')
    
  print(ggraph(g, layout = 'circle') + 
          geom_edge_fan(edge_linetype = 3, color = 'dark blue', alpha = 0.25) + 
          geom_node_point(color = 'dark red', size = 2, alpha = 0.75) + 
          theme_graph(base_family = 'Helvetica') +
          labs(title = 'An Example of Barabasi-Albert Model',
               subtitle = glue('with ', n, ' Nodes, ', m, ' Edges Added at Each Step, \nand the Power of Preferential Attachment of ', power)))
  
  writeLines('\n')

  suppressMessages(df <- bind_cols(enframe(eccentricity(g)), 
                                   enframe(betweenness(g)), 
                                   enframe(degree(g)), 
                                   enframe(transitivity(g, type = c('local')))))
  
  df <- df %>% select(name...1, value...2, value...4, value...6, value...8)

  names(df) <- c('name', 'eccentricity', 'betweenness', 'degree', 'clustering')
  
  DF2 <- df %>% 
    summarize(power = power,
              avg_deg = mean(degree), 
              delta = max(degree), 
              diam = max(eccentricity),
              radius = min(eccentricity),
              avg_cc = mean(clustering, na.rm = TRUE), 
              exponent = fit_power_law(degree, implementation = 'plfit')$alpha)
  
  DF <- rbind(DF, DF2)
  
}

DF
```

```{r}
set.seed(42)

n <- 1000

for (power in c(0, 0.5, 1, 1.5, 2, 2.5, 3)) {
  
  m <- 8
  
  g <- sample_pa(n, m = m, power = power, directed = FALSE, algorithm = 'psumtree')

  df <- enframe(degree(g))
  
  df <- df %>% select(name, value)

  names(df) <- c('name', 'degree')
  
  exponent <- fit_power_law(df$degree, implementation = 'plfit')$alpha
  
  df2 <- logbin(df$degree, n = 100)
  
  total2 <- sum(df2$bin_value)

  print(df2 %>% 
          filter(bin_count != 0) %>% 
          ggplot(aes(x = bin_midpoint, y = bin_value / total2)) + 
          geom_point(color = 'blue', size = .5) + 
          geom_line(data = df, 
                    aes(x = degree, y = 2 * m ** (exponent - 1) * (degree) ** (-exponent)), 
                    color = 'red') + 
          geom_line(data = df, 
                    aes(x = degree, y = 2 * m ** 2 * (degree) **(-3)), 
                    color = 'yellow') + 
          scale_x_log10() + 
          scale_y_log10() + 
          labs(y = TeX('$p_k$'), x = 'k') +
          labs(title = 'Log-Log Scale, Linear Binning for Degree Distribution',
               subtitle = glue('with Fitted Power Law Curves (Eq. 5.9: Red; Eq. 5.15: Yellow) in a Barabasi-Albert Graph \nwith ', n, ' Nodes, ',  m, ' Edges Added at Each Step, \nand the Power of Preferential Attachment of ', power)))
  
  writeLines('\n')

}
```

# References

- Albert-Laszlo Barabasi, *Network Science*, Cambridge University Press, 2016. *Network Science* is available online at http://networksciencebook.com/ under the following license: "This book's text and illustrations are licensed under a [Creative Commons Attribution-NonCommercial 3.0 Unported License](https://creativecommons.org/licenses/by-nc/3.0/)."
